Prestige bias in cultural evolutionary dynamics

### 2023-11-12 ms/ms.tex
continue revision for rev2

Following \cref{eq:ourF}, the probability that the $i$-th copier chose role-model $j$ is given by the \emph{influence} of role-model $j$ in the eyes of copier $i$,
\begin{equation}\label{eq:recPrestige}
Pr(S_{i,j}=1 \mid S_{1,j},S_{2,j},...,S_{i-1,j}) = G_{i,j} \;.
\end{equation}
The influence $G_{i,j}$ of role-model~$j$ in the eyes of copier~$i$ is determined by success--the estimated biased trait value $\beta(A_{i,j})$--and prestige--the number of copiers that chose role-model $j$ before copier $i$, $K_{i-1,j}$, replacing \cref{eq:boydG} with
\begin{equation}\label{eq:prestige}
G_{i,j} = \frac{\alpha_{ij} \cdot \beta(A_{i,j}) + (1-\alpha_{ij}) \cdot K_{i-1,j}}{W_i} \;,
\end{equation}
where $W_i$ is a normalizing factor that sums the numerator over all role-models ($1\le j \le N)$ to ensure $\sum_{j=1}^{N}{G_{i,j}}=1$.

Here, the bias weight $\alpha_{i,j}$ determines the relative weighting of success and prestige bias. It is a characteristic of the interaction of role-model~$j$ with copier $i$ that determines the relative significance of direct success vs. prestige in the role-model's overall influence in the eyes of the copier.
Different individuals may evaluate the importance of success and prestige differently. Additionally, we assume each role-model displays its prestige and success individually. For example, individuals with more followers but lacking skill may emphasize the number of their followers rather than their skill (i.e., have lower $\alpha_{i,j}$ value).
Finally, the trait of role-model $j$ estimated by copier $i$, $A_{i,j}$, remains as in \cref{eq:relativeTrait}.

\subsection*{Dichotomous trait}
We introduce a simplified version of the above model where the trait has only two phenotypes: an optimal phenotype $\hat{A}$ and a sub-optimal phenotype $A$. 
All role-models with the same phenotype will contribute to the probability that that phenotype is transmitted and thus influence is determined by the number of copiers that have already chosen a role-model with either phenotype.
In addition, for simplicity and for easier mathematical analysis, we assume $\alpha$ is homogeneous, {which entails} exchangeability between role-models.
Therefore, the probability {that} the $i$-th copier {copies} phenotype $A$ is
\begin{equation}\label{eq:binary-model}
G_{i,A} = \frac{(N-X)\alpha'\beta(A) + K_{i-1,A}}{i-1 + (N-X)\alpha'\beta(A) + \alpha'X} \;,
\end{equation}
where $X$ is the number of role-models with trait $\hat{A}$; $K_{i,A}$ is the number of copiers that already chose $A$ when copier $i$ chooses a role-model; assuming that $\beta(\hat{A})=1$; and defining $\alpha'= \frac{\alpha}{1-\alpha}$ as the odds ratio of the bias weight.
Complementing this, the probability of the $i$-th copier to copy phenotype $\hat{A}$ is $G_{i, \hat{A}} = 1-G_{i,A}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
Our models are defined by two nested stochastic processes. Change over multiple generations is described by the {dynamics of the} phenotype distribution at each generation, $\{\vec{A}(t)\}_t$, see \cref{eq:transmission}. The transition from one generation to the next is described by the number of copiers each role-model has after $i$ copiers have chosen a role-model, $\{\vec{K}_i\}_{i=1}^N$, see \cref{eq:process}.
We emphasize that the models are nested: $\vec{A}(t+1)$ can be computed from $\vec{A}(t)$ by evaluating $\vec{K}_{N}$, where $K_{N,j}$ is the number of copiers that chose role-model $j$ after all copiers chose a role model.
However, the former requires iterating over \cref{eq:recurrence,eq:recPrestige}.
Thus, we sought to find an equivalent stochastic process that has the same joint distribution as $\vec{K}_{N}$. 
We found two approximations for the distribution of $\vec{K}_{N}$, summarized here and explained in detail below.
In both we assume that the bias weight is either completely homogeneous, $\alpha_{i,j}=\alpha$, or that $\alpha_{i,j}=\alpha_j$ is a bias of role-model $j$ only, meaning it does not vary between copiers.
Note that these approximations apply for both the and the continuous trait (\cref{eq:prestige}) and the dichotomous trait (\cref{eq:binary-model}) models.

@@

\begin{result}[Generalized binomial distribution approximation]\label{res:GBD}
The number of copiers of role-model $j$ after $i$ copiers have chosen a role-model follows the generalized binomial distribution, $$K_{i,j} \sim \textit{GenBin}(i,\alpha_j\cdot\beta(A_j+e),\theta)$$ if $e_i=e$ for all role-models $i=1, \ldots, N$, and $\theta$ is the correlation between successive role-model choices.
\end{result}

@@

\begin{result}[P{\'{o}}lya urn model]\label{result:polya}
The role-model choice process, $\big\{\vec{K}_i\big\}_{i=1}^N$, is equivalent to a \emph{P\'{o}lya urn model} if both trait estimation error and bias weight are uniform in the population, $e=e_j$ and $\alpha=\alpha_j$ for all $j=1,\ldots,N$.
\end{result}

\begin{proof} 
{Write} $\alpha'=\frac{\alpha}{1-\alpha}$ as the bias weight ratio, and $A'_j=A_j+e$. From \cref{eq:prestige} and because $\sum_{j=1}^{N}{K_{i,j}}=i$, we have
\begin{equation}\label{eq:copier_choose}
G_{i,j} = 
\frac{\alpha'\beta(A'_j) + K_{i-1,j}}{\sum\limits_{m=1}^{N}{\left[ \alpha'\beta(A'_m) + K_{i-1,m}\right]}}
 =\frac{\alpha'\beta(A'_j) + K_{i-1,j}}{i-1 + \sum\limits_{m=1}^{N}\alpha'\beta(A'_m)} \;.
\end{equation}
Substituting $M=N$, $o_j = \alpha'\beta(A'_j)$, and $w_{i,j} = K_{i,j}$ in \cref{eq:polya} gives \cref{eq:copier_choose}, thus completing the proof.
\end{proof}

@@

For simplicity, we use the dichotomous-trait model, and we do not include role-model estimation error in this analysis, i.e $e_i=0$ for every copier $i$.
As shown above, transmission in our model is approximately Dirichlet-Multinomial distributed (\Cref{cor:dirichlet} and \cref{eq:copier_choose}).

@@

\begin{figure}[H]
    \includegraphics[width=\linewidth]{ch_env.pdf}
  \caption{\textbf{Fixation probability in a changing environment.}
\textbf{(A)} Fixation probability decreases with {the} success-bias weight (bottom x-axis) and effective population size (top x-axis). The approximation (blue; \cref{eq:ch_env}) agrees with simulation results (orange). 
\textbf{(B)} Fixation probability increases with the success coefficient, $\beta$.
When success bias is large ($1-\beta > 0.1$),  
simulation results (orange) are underestimated by the changing environment approximation (blue; \cref{eq:ch_env}). With even larger success bias ($1-\beta > 0.35$), even the constant environment approximation (green; \cref{eq:kimura_p}) slightly underestimates simulation results, likely because the diffusion equation approximation assumes weak "selection" .
(\textbf{C,D}) The approximation (blue) is robust to changes in environmental cycle length, as it agrees with simulations (orange) for different sizes of the changing environment cycle, where $k$ and $l$ are the number of generations each trait value is under success bias. 
When $k>l$, the approximation and the simulations are both very close to the constant environment approximation (green), because the more generations the rare phenotype is favored, the more similar it is to the constant environment model, where it is always favored by the success bias.
Markers show average of $10,000$ simulations, error bars show 75\% (panels A, C, and D) and 95\% (panel B) confidence intervals.
  Here, population size, $N=1,000$; phenotype values, $\hat{A}=1$ with $A=0.9$, $k=20$, $l=80$ (panels A and B) and $A=0.8$ (panels C and D); In panel A, the success coefficient is $1-\beta=s=0.005$; In panels B, C, and D, the success-bias weight is $\alpha=0.1$.
  }
  \label{fig:ch_env_alpha_beta}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Optimal success-bias weight}
% TODO edit the next sentence.
We ran simulations of the role-model choice process during a single generation in which every copier evaluates its own optimal success-bias weight, $\alpha^*$, which minimizes the expected squared error between the chosen trait value and the optimal trait value $\hat{A}$,
\begin{equation}\label{eq:optimal_alpha}
\alpha^* = \emph{argmin} \sum_{j=1}^N\frac{\alpha A_j + (1-\alpha) K_j}{\sum_{l=1}^N\alpha A_l + (1-\alpha) K_l} (\hat{A}-A_j)^2 \;,
\end{equation}
where $A_j$ is the trait of role-model $j$ and $K_j$ the number of copiers that already chose role-model~$j$.
Simply put, each copier estimates what success-bias weight $\alpha^*$ will result in copying a trait that is most similar to the ideal trait value $\hat{A}$.
Indeed, if the trait value is correlated with fitness, the optimal success-bias weight would increase the fitness of individuals. However, here we ignore the effects of natural selection, focusing instead on selectively neutral traits.

We find that when copiers choose their success-bias weight according to \cref{eq:optimal_alpha}, it decreases with the number of copiers that have already chosen a role-model~(\Cref{fig:influence_advantage}).
Moreover, their estimation error is much lower compared to a constant success-bias weight, which gives roughly the same high estimation error to all copiers (compare \Cref{fig:influence_advantage}B and C): in this example, the optimal success-bias weight gives an estimation error (difference between the chosen and optimal trait) converges to $0.046$, whereas a constant success-bias weight gives values $>0.74$.

\begin{figure}[h]
    \includegraphics[width=\linewidth]{choose_bias.pdf}
  \caption{
  \textbf{Advantage of an optimal success-bias weight.}
  Both success-bias weight $\alpha$ (\textbf{A}) and estimation error (\textbf{B}) decrease during the role-model choosing process (within a single generation), demonstrating that prestige becomes more favored by copiers as more copiers have made their choice.
However, when $\alpha$ is homogeneous (\textbf{C}), the mean estimation error doesn't decrease, regardless of $\alpha$ or $\eta$.
The mean estimation error in the homogeneous $\alpha$ model is larger by a factor of $10$ than the optimal $\alpha$ model.
Here "Copier" of the x-axis is the index of the choosing copier, population size $N=200$; estimation error is normally distributed $e \sim N(0,\eta^2)$ with standard deviation $\eta=$0.0001 (blue), 0.001 (orange), 0.01 (green), plots are average of $300$ simulations.}	
  \label{fig:influence_advantage}
\end{figure}

>>>>

Following \cref{eq:ourF}, the probability that the $i$-th copier chose role-model $j$ is given by the \emph{influence} of role-model $j$ in the eyes of copier $i$,
\begin{equation}\label{eq:recPrestige}
Pr(S_{i,j}=1 \mid S_{1,j},S_{2,j},...,S_{i-1,j}) = G_{i,j} \;.
\end{equation}
The influence $G_{i,j}$ of role-model~$j$ in the eyes of copier~$i$ is determined by success--the estimated biased trait value $\beta(A_{i,j})$--and prestige--the number of copiers that chose role-model $j$ before copier $i$, $K_{i-1,j}$, replacing \cref{eq:boydG} with
\begin{equation}\label{eq:prestige}
G_{i,j} = \frac{\alpha_{ij} \cdot \beta(A_{i,j}) + (1-\alpha_{ij}) \cdot K_{i-1,j}}{W_i} \;,
\end{equation}
where $W_i$ is a normalizing factor that sums the numerator over all role-models ($1\le j \le N)$ to ensure $\sum_{j=1}^{N}{G_{i,j}}=1$.
Here, the success-bias weight $\alpha_{i,j}$ determines the relative weighting of success and prestige bias. It is a characteristic of the interaction of role-model~$j$ with copier $i$ that determines the relative significance of direct success vs. prestige in the role-model's overall influence in the eyes of the copier.
Different individuals may evaluate the importance of success and prestige differently. Additionally, we assume each role-model displays its prestige and success individually. For example, individuals with more followers but lacking skill may emphasize the number of their followers rather than their skill (i.e., have lower $\alpha_{i,j}$ value).
Finally, the trait of role-model $j$ estimated by copier $i$, $A_{i,j}$, remains as in \cref{eq:relativeTrait}.

\subsection*{Dichotomous trait}
We introduce a simplified version of the above model where the trait has only two phenotypes: an optimal phenotype $\hat{A}$ and a sub-optimal phenotype $A$. 
All role-models with the same phenotype will contribute to the probability that that phenotype is transmitted and thus influence is determined by the number of copiers that have already chosen a role-model with either phenotype.
In addition, for simplicity and for easier mathematical analysis, we assume $\alpha$ is homogeneous and constant ($\alpha_{i,j}=\alpha$), which entails exchangeability between role-models.
Therefore, the probability that the $i$-th copier {copies} phenotype $A$ is
\begin{equation}\label{eq:binary-model}
%G_{i,A} = \frac{(N-X)\alpha'\beta(A) + K_{i-1,A}}{i-1 + (N-X)\alpha'\beta(A) + \alpha'X} \;,
G_{i,A} = \frac{\alpha \cdot (N-X) \beta(A) + (1-\alpha) \cdot K_{i-1,A}}{\alpha \cdot (N-X) \beta(A) + \alpha \cdot X + (1-\alpha)\cdot(i-1)} \;,
\end{equation}
where $X$ is the number of role-models with trait $\hat{A}$ and $K_{i-1,A}$ is the number of copiers that already chose $A$ when copier $i$ chooses a role-model, and assuming that $\beta(\hat{A})=1$ (thus the term $\alpha X$ in the denominator).
Complementing this, the probability of the $i$-th copier to copy phenotype $\hat{A}$ is $G_{i, \hat{A}} = 1-G_{i,A}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
Our models are defined by two nested stochastic processes. Change over multiple generations is described by the {dynamics of the} phenotype distribution at each generation, $\{\vec{A}(t)\}_t$, see \cref{eq:transmission}. The transition from one generation to the next is described by the number of copiers each role-model has after $i$ copiers have chosen a role-model, $\{\vec{K}_i\}_{i=1}^N$, see \cref{eq:process}.
We emphasize that the models are nested: $\vec{A}(t+1)$ can be computed from $\vec{A}(t)$ by evaluating $\vec{K}_{N}$, where $K_{N,j}$ is the number of copiers that chose role-model $j$ after all copiers chose a role model.
However, the former requires iterating over \cref{eq:recurrence,eq:recPrestige}.
Thus, we sought to find an equivalent stochastic process that has the same joint distribution as $\vec{K}_{N}$. 
We found two approximations for the distribution of $\vec{K}_{N}$, summarized here and explained in detail below.
In both approximations we assume that the success-bias weight is either completely homogeneous, $\alpha_{i,j}=\alpha$, or that $\alpha_{i,j}=\alpha_j$ is a characteristic of role-model $j$ that does not vary between copiers.
Note that these approximations apply for both the and the continuous trait (\cref{eq:prestige}) and the dichotomous trait (\cref{eq:binary-model}) models.

@@

\begin{result}[Generalized binomial distribution approximation]\label{res:GBD}
The number of copiers of role-model $j$ after $i$ copiers have chosen a role-model follows the generalized binomial distribution, 
$$
K_{i,j} \sim \textit{GenBin}(i,\alpha_j\cdot\beta(A_j+e),\theta)
$$
if $e_i=e$ for all copiers $i=1, \ldots, N$; the success-bias weight only depends on the role-model and not the copier, i.e., $\alpha_{i,j}=\alpha_j$  for all $i=1, \ldots, N$; and $\theta$ is the correlation between successive role-model choices.
\end{result}

@@

\begin{result}[P{\'{o}}lya urn model]\label{result:polya}
The role-model choice process, $\big\{\vec{K}_i\big\}_{i=1}^N$, is equivalent to a \emph{P\'{o}lya urn model} if both the trait estimation error and the success-bias weight are constant in the population, $e_i=e$ for all $i=1, \ldots, N$ and $\alpha_{i,j}=\alpha$  for all $i,j=1, \ldots, N$.
\end{result}

\begin{proof} 
{Write} $\alpha'=\frac{\alpha}{1-\alpha}$ as the success-bias weight ratio, and $A'_j=A_j+e$. From \cref{eq:prestige} and because $\sum_{j=1}^{N}{K_{i,j}}=i$, we have
\begin{equation}\label{eq:copier_choose}
G_{i,j} = 
\frac{\alpha'\beta(A'_j) + K_{i-1,j}}{\sum\limits_{m=1}^{N}{\left[ \alpha'\beta(A'_m) + K_{i-1,m}\right]}}
 =\frac{\alpha'\beta(A'_j) + K_{i-1,j}}{i-1 + \sum\limits_{m=1}^{N}\alpha'\beta(A'_m)} \;.
\end{equation}
Substituting $M=N$, $o_j = \alpha'\beta(A'_j)$, and $w_{i,j} = K_{i,j}$ in \cref{eq:polya} gives \cref{eq:copier_choose}, thus completing the proof.
\end{proof}

@@

For simplicity, we use the dichotomous-trait model, which also assumes a constant success-bias weight $\alpha_{i,j}=\alpha$, and we do not include trait estimation error in this analysis, i.e., $e_i=0$.
As shown above, transmission in our model is approximately Dirichlet-Multinomial distributed (\Cref{cor:dirichlet} and \cref{eq:copier_choose}).

@@

\begin{figure}[H]
    \includegraphics[width=\linewidth]{ch_env.pdf}
  \caption{\textbf{Fixation probability in a changing environment.}
\textbf{(A)} Fixation probability decreases with {the} success-bias weight (bottom x-axis) and effective population size (top x-axis). The approximation (blue; \cref{eq:ch_env}) agrees with simulation results (orange). 
\textbf{(B)} Fixation probability increases with the success coefficient, $\beta$.
When success bias is large ($1-\beta > 0.1$),  
simulation results (orange) are underestimated by the changing environment approximation (blue; \cref{eq:ch_env}). With even larger success bias ($1-\beta > 0.35$), even the constant environment approximation (green; \cref{eq:kimura_p}) slightly underestimates simulation results, likely because the diffusion equation approximation assumes weak "selection" .
(\textbf{C}) The approximation (blue) is robust to changes in environmental cycle length, as it agrees with simulations (orange) for different sizes of the changing environment cycle, where $k$ and $l$ are the number of generations each trait value is under success bias. 
When $k>l$, the approximation and the simulations are both very close to the constant environment approximation (green), because the more generations the rare phenotype is favored, the more similar it is to the constant environment model, where it is always favored by the success bias.
Markers show average of $10,000$ simulations, error bars show 75\% (panels A and C) and 95\% (panel B) confidence intervals.
Here, population size, $N=1,000$; phenotype values, $\hat{A}=1$ with $A=0.9$, $k=20$, $l=80$ (panels A and B) and $A=0.8$ (panel C); In panel A, the success coefficient is $1-\beta=s=0.005$; In panels B and C, the success-bias weight is $\alpha=0.1$.
  }
  \label{fig:ch_env_alpha_beta}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Optimal success-bias weight}
In results 2-6, we assumed that $\alpha$ is homogeneous in the population and constant, that is, it does not depend on any specific context.
Next, we examined what happens in the continuous-trait model if the $i$-th copier evaluates its own optimal success-bias weight, $\alpha^*_i$, which minimizes the expected squared error between the chosen trait value and the optimal trait value $\hat{A}$,
\begin{equation}\label{eq:optimal_alpha}
\alpha^*_i = \emph{argmin} \sum_{j=1}^N\frac{\alpha A_j + (1-\alpha) K_{i-1,j}}{\sum_{l=1}^N\alpha A_l + (1-\alpha) K_{i-1,l}} (\hat{A}-A_j)^2 \;,
\end{equation}
where $A_j$ is the trait of role-model $j$ and $K_{i-1,j}$ is the number of copiers that already chose role-model~$j$ by the time the $i$-the copier chooses a role-model.
Simply put, each copier $i$ estimates what success-bias weight $\alpha^*_i$ will result in copying a trait that is most similar to the ideal trait value $\hat{A}$.
Indeed, if the trait value is correlated with fitness, the optimal success-bias weight would increase the fitness of individuals. However, here we ignore the effects of natural selection, focusing instead on selectively neutral traits.

We find that when copiers choose their success-bias weight according to \cref{eq:optimal_alpha}, it quickly decreases with the number of copiers that have already chosen a role-model and then stays at what appears to be an equilibrium~(\Cref{fig:influence_advantage}).
Moreover, their estimation error is much lower compared to a constant success-bias weight, which gives roughly the same high estimation error to all copiers (compare \Cref{fig:influence_advantage}B and C): in this example, the optimal success-bias weight gives an estimation error (difference between the chosen and optimal trait) converges to $0.046$, whereas a constant success-bias weight gives values $>0.74$.

\begin{figure}[h]
    \includegraphics[width=\linewidth]{choose_bias.pdf}
  \caption{
  \textbf{Advantage of an optimal success-bias weight.}
  Both success-bias weight $\alpha$ (\textbf{A}) and estimation error (\textbf{B}) decrease during the role-model choosing process (within a single generation), demonstrating that prestige becomes more favored by copiers as more copiers have made their choice.
However, when $\alpha$ is homogeneous (\textbf{C}), the mean estimation error does not decrease, regardless of $\alpha$ or $\eta$.
The mean estimation error in the homogeneous $\alpha$ model is larger by a factor of $10$ than the optimal $\alpha$ model.
Here "Copier" of the x-axis is the index of the choosing copier, population size $N=200$; estimation error is normally distributed $e \sim N(0,\eta^2)$ with standard deviation $\eta=$0.0001 (blue), 0.001 (orange), 0.01 (green), markers are average of $300$ simulations.
% lines are also 300 simulations? why not markers? 
% what are error bars in C? CI 99%
% C x axis should say "Constant success-bias weight"
% A B x axis should say "Copier $i$; A y axis should be "Optimal success-bias weight, \alpha^*_i$.
}	
  \label{fig:influence_advantage}
\end{figure}
